Arrows for Parallel Computations

### 2017-07-21 paper/content/conclusion.tex
intro + conclusion: a first version

\section{Conclusion}
\label{sec:conclusion}
Arrows are a generic concept that allows for powerful composition combinators. To our knowledge we are the first ones to represent parallel computation with arrows.\olcomment{that strange arrows-based robot interaction paper from 1993 or so! clearly discuss in related work} \done

Arrows turn out to be a useful tool for composing in parallel programs. We do not have to introduce new monadic types that wrap the computation. Instead use arrows just like regular sequential pure functions. 
%
This work features multiple parallel backends: the already available parallel Haskell flavours. Parallel Arrows feature an implementation of the |ArrowParallel| interface for Multicore Haskell, |Par| Monad, and Eden. With our approach parallel programs can be ported across these flavours with little to no effort.
%
%
Performance-wise, Parallel Arrows are on par with existing parallel Haskells, as they do not introduce any notable overhead.

The strictness problems of the |Par| Monad backend materializes with |ArrowLoop|. Additional work is required to ensure correct behaviour of the |Par| Monad in this context.

\mbcomment{mention ArrowLoop in Torus and Ring chapters}
\olcomment{Parrows + accelerate = love? Metion port to Frege.}

@@

Our PArrows DSL can be expanded to futher parallel Haskells. More specifically we target HdpH \cite{Maier:2014:HDS:2775050.2633363}, a modern distributed Haskell that would benefit from our Arrows notation. More Future-aware versions of Arrow combinators can be defined and existing can be further improved. We would look into more transparency of the DSL, it should basically infuse as little overhead as possible.

We are looking into more experiences with seamless porting of parallel PArrow-based programs across the backends.
Of course, we are working on expanding both our skeleton library and the number of skeleton-based parallel programs that use our DSL to be portable across flavours of parallel Haskells.
It would also be interesting to see a hybrid of PArrows and Accelerate.
Ports of our approach to other languages like Frege or Java directly are in an early development stage.

>>>>

\section{Conclusion}
\label{sec:conclusion}
Arrows are a generic concept that allows for powerful composition
combinators. To our knowledge we are the first ones to represent
\emph{parallel} computation with Arrows.
% \olcomment{that strange arrows-based robot interaction paper from 1993 or so! clearly discuss in related work} \done
%
Arrows turn out to be a useful tool for composing in parallel
programs. We do not have to introduce new monadic types that wrap the
computation. Instead, we use Arrows in the same manner one uses sequential pure functions. 
%
This work features multiple parallel backends: the already available parallel Haskell flavours. Parallel Arrows, as presented here, feature an implementation of the |ArrowParallel| typeclass for Multicore Haskell, |Par| Monad, and Eden. With our approach parallel programs can be ported across these flavours with little to no effort. It is quite straightforward to add further backends. 
%
%
Performance-wise, Parallel Arrows are on par with existing parallel Haskells, as they do not introduce any notable overhead.
%
The benefit is, however, the greatly increased portability of parallel
programs.

\mbcomment{mention ArrowLoop in Torus and Ring chapters}
%\olcomment{Parrows + accelerate = love?} \olcomment{Metion port to Frege.}

@@

Our PArrows DSL can be expanded to further parallel Haskells. More specifically we target HdpH \cite{Maier:2014:HDS:2775050.2633363} for this future extension. HdpH is a modern distributed Haskell that would benefit from our Arrows notation. Further Future-aware versions of Arrow combinators can be defined. Existing combinators could also be improved. We would look into more transparency of our DSL,  basically it should infuse as little overhead as possible.

More experiences with seamless porting of parallel PArrows-based programs across the backends are welcome.
Of course, we are working ourselves on expanding both our skeleton library and the number of skeleton-based parallel programs that use our DSL to be portable across flavours of parallel Haskells.
It would also be interesting to see a hybrid of PArrows and Accelerate \cite{McDonell:2015:TRC:2887747.2804313}.
Ports of our approach to other languages like Frege or Java directly are in an early development stage.

### 2017-07-21 paper/content/motivation.tex
intro + conclusion: a first version

\section{Introduction}
\label{sec:introduction}
\olcomment{todo, reuse 5.5, and more}

\mbcomment{
Haskell is Spielwiese für Parallelität; verschiedene Ansätze (Par, Multicore, Eden); Orthogonale Ansätze; Verwenden höchstens eine Monade - manchmal auch nur intern; Wir wollen Parallelität mit Arrows abbilden, was noch niemand gemacht hat;
Statt einer eigenen Implementierung definieren wir ein "shallow embedded DSL" (ACHTUNG, ist das der richtige Name? effektiv API);
Umsetzung mit verschiedenen parallelen Haskells; We tame the zoo of parallel Haskells und vergewissern uns dass es nicht viel Overhead bringt
}

blablabla arrows, parallel, haskell.

\paragraph{Contribution}

\olcomment{HIT HERE REALLY STRONG}

%\subsection{Impact of parallel Arrows}
%\olcomment{move this to Contributions in the front or something}
\mbcomment{different, how?}
We wrap parallel Haskells inside of our |ArrowParallel| interface, but why do we aim to abstract parallelism this way and what does this approach do better than the other parallel Haskells?
\begin{itemize}
	\item \textbf{Arrow DSL benefits}:
	With the |ArrowParallel| typeclass we do not lose any benefits of using arrows as |parEvalN| is a yet another Arrow combinator. The resulting Arrow can be used in the same way a potential serial version could be used. This is a big advantage of this approach, especially compared to the Monad solutions as we do not introduce any new types. We can just \enquote{plug} in parallel parts into sequential Arrow-based programs without having to change anything.
	\item \textbf{Abstraction}:
	With the |ArrowParallel| typeclass, we abstract all parallel implementation logic away from the business logic. This leaves us in the beautiful situation of being able to write our code against the interface of the typeclass without being bound to any parallel Haskell. So as an example, during development, we can run the program in a simple GHC-compiled variant and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the actual |ArrowParallel| instance.
\end{itemize}

>>>>

\section{Introduction}
\label{sec:introduction}
% \olcomment{todo, reuse 5.5, and more}
%
% \mbcomment{
% Haskell is Spielwiese für Parallelität; verschiedene Ansätze (Par, Multicore, Eden); Orthogonale Ansätze; Verwenden höchstens eine Monade - manchmal auch nur intern; Wir wollen Parallelität mit Arrows abbilden, was noch niemand gemacht hat;
% Statt einer eigenen Implementierung definieren wir ein "shallow embedded DSL" (ACHTUNG, ist das der richtige Name? effektiv API);
% Umsetzung mit verschiedenen parallelen Haskells; We tame the zoo of parallel Haskells und vergewissern uns dass es nicht viel Overhead bringt
% }
%
% blablabla arrows, parallel, haskell.
%
% \olcomment{An attempt ---->}
%%%%%

Pure functional programming languages have been used
since longer time as a ``sand box'' for testing novel parallel
programming approaches. Reasons for this include referential
transparency and the ease of introducing new sophisticated language
concepts. Specially for Haskell (on which this paper focuses), various
approaches reached maturity. The definite parallel Haskell does not
exist, there are multiple parallel Haskells. We regard here in-depth
the Multicore Haskell (a SMP implementation of GpH language), the
|Par| Monad, and Eden, a distributed memory parallel Haskell. These
languages represent orthogonal approaches. Some use a monad, even if
only for the internal representation. Some introduce additional
language constructs. 
%In this paper we introduce a notion of parallel
%computations using Arrows.
We use here Arrows to represent parallel computations. This has not
been done before.

%Our Arrows-based interface is a high-level one.
We provide an Arrows-based typeclass and various implementations.
Instead of 
introducing a new low-level parallel backend in order to implement our
Arrows-based interface, we define a shallow-embedded DSL for Arrows. This DSL
is defined as a common interface and varying implementations in
existing parallel Haskells: Multicore Haskell, |Par| Monad, and
Eden. Thus, we not only define a parallel programming interface in a
novel manner. We tame the zoo of parallel Haskells. We provide a
common, very low-penalty programming interface that allows to switch
the parallel Haskell backends at will. Further backends are viable, too.

\paragraph{Contribution}
%
%\olcomment{HIT HERE REALLY STRONG}
%
%\subsection{Impact of parallel Arrows}
%\olcomment{move this to Contributions in the front or something}
\mbcomment{different, how?}
We wrap parallel Haskells inside of our |ArrowParallel| interface, but
% why do we aim to abstract parallelism this way and what does this
% approach do better than the other parallel Haskells?
 is such a parallelism abstraction of benefit and to what extent does
 it improve existing approaches?
\begin{itemize}
	\item \textbf{Arrow DSL benefits}:
	With the |ArrowParallel| typeclass we do not lose any benefits of using arrows as |parEvalN| is a yet another Arrow combinator. The resulting Arrow can be used in the same way a potential serial version could be used. This is a big advantage of this approach, especially compared to the Monad solutions as we do not introduce any new types. We can just \enquote{plug} in parallel parts into sequential Arrow-based programs without having to change anything.
	\item \textbf{Abstraction}:
	With the |ArrowParallel| typeclass, we abstract all parallel implementation logic away from the business logic. This leaves us in the beautiful situation of being able to write our code against the interface of the typeclass without being bound to any parallel Haskell. So as an example, during development, we can run the program in a simple GHC-compiled variant and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the current |ArrowParallel| instance.
\end{itemize}
